Parallel Sorting of Roughly-Sorted Sequences\\CSCI 5172 Fall '16 Project

### 2016-12-10 tex/paper.tex
adding figure PNG, adding credit

The \textit{comparison} sorts are an indispensable family of sorting algorithms containing such classics as Heapsort,
  Mergesort and Quicksort.
True to its name, algorithms in this family determine how to reorder a sequence by comparing its elements against each other
  according to some natural or explicitly-provided ordering;
  for the sake of simplicity, we shall only consider the sorting of integral elements in nondecreasing order, backed by
  random-access array storage\footnote{Sequential-access storage demands distinct and divergent design considerations.}.
A well-established result is the ``linearithmic'' $\Omega(n \lg n)$ runtime lower bound for the general problem of sorting
  an array of $n$ elements by comparison; algorithms such as Mergesort achieve $\Theta(n \lg n)$ runtime (and are thus
  asymptotically optimal) whereas Quicksort, while usually outperforming Mergesort over arrays that fit in memory,
  has quadratic-runtime pathological cases.

@@

@@

@@

@@

Regrettably, Thrust has yet to provide an \texttt{nth_element} selection algorithm implementation, so we replaced it with calls
  to Thrust's \texttt{sort} algorithm, which performs a ``state-of-the-art'', highly-optimized radix sort
  implementation when invoked over arrays of primitive integer elements\cite{thrustradix, leischner09}.
Radix sort is a generalized counting sort that runs in linear time when operating over fixed-size array elements.
While radix sort requires additional storage (unlike Introselect, which was used in the sequential implementation), this
  additional storage ought to be acceptable as the segments we wish to partition ought to be relatively small.

Each Thrust \texttt{sort} call issues at least one kernel launch; we attempted to run radix sorts over the entire array in
  parallel by establishing a fixed number of CUDA \textit{streams}.
Whereas kernels launched through the same stream are executed in order, up to 16 kernels launched through distinct streams can be
  run concurrently on the GPU, available resources permitting\cite{cuda}.
We established a fixed number of CUDA streams and launched each \texttt{sort} call through them in a circular, round-robin
  fashion.

@@

We implemented this approach using a CUDA kernel where each thread in the launch is responsible for sorting exactly one segment
  from step 1 or 2; the steps are divided into separate kernel launches, each given an offset from which begin segmenting the
  array.
While the threads in each launch run concurrently\footnote{For large arrays, the GPU can't host every thread simultaneously and
  must sequentially execute the thread blocks in batches.}, each thread sorts its designated segment sequentially.
In contrast to \supercite{altman90}, we used Thrust's radix sort instead of its Mergesort implementation in order to leverage
  the linear runtime of the former.
While this approach to the parallel Roughsort was far more successful than the previous, it still struggled against
  its competitors (see Section 9).

@@

\section{Implementation Performance Analysis}
We tested our sequential and parallel Roughsort implementations against sequential and parallel Mergesort (as implemented by
  the C\texttt{++} STL and by CUDA Thrust, respectively) as well as against our simple, sequential Bubblesort implementation.
We chose Mergesort since it's a classic sort algorithm that also scales very well with parallel execution, capable
  of achieving a speedup of at least $\Theta(n / \lg^2 n)$ when given unlimited processors for a runtime of
  $\Theta(\lg^3 n)$\cite{clrs}.
While Bubblesort has a quadratic runtime upper bound, its lightweight, linear lower bound emerges when given nearly-sorted
  sequences and provides a means of comparing our implementation's performance for small values of $k$.

@@

Figures 5 and 6 fix $n$ and vary $k$, yielding similar results.
For arrays of these lengths, the parallel Roughsort's crossover point lies somewhere around the interval $20 \leq k \leq 30$,
  where it manages to best the three sequential algorithms.
An invariant among all four graphs is the utter dominance of the parallel Mergesort implementation provided by Thrust,
  whose runtimes never depart from the $x$-axis.

@@

\clearpage
\section{Explanation of Results}

\section{Conclusion and Further Research}
An implementation of Cole's parallel median algorithm might enable a more efficient parallel Roughsort\cite{cole85}.

>>>>

The \textit{comparison} sorts are an indispensable family of sorting algorithms containing such classics as Heapsort,
  Mergesort and Quicksort.
True to its name, algorithms in this family determine how to reorder a sequence by comparing its elements against each other
  according to some natural or explicitly-provided ordering;
  for the sake of simplicity, we shall only consider the sorting of integral elements in non-decreasing order, backed by
  random-access array storage\footnote{Sequential-access storage demands distinct and divergent design considerations.}.
A well-established result is the ``linearithmic'' $\Omega(n \lg n)$ runtime lower bound for the general problem of sorting
  an array of $n$ elements by comparison; algorithms such as Mergesort achieve $\Theta(n \lg n)$ runtime (and are thus
  asymptotically optimal) whereas Quicksort, while usually outperforming Mergesort over arrays that fit in memory,
  has quadratic-runtime pathological cases.

@@

  Parallel implementations of the Roughsort and radius finding algorithms are also addressed.  The aforementioned algorithms 
  assume a specific underlying memory in achieving the subsequent parallel implementation runtimes: the concurrent-read 
  concurrent-write parallel random access model (CRCW PRAM).  Shiloach and Vishkin elaborate on this memory model, providing
  that it is a purely theoretical model which, \textit{inter alia}, meets the following criteria: a shared, global memory of
  size $m \geq p$ is available to $p$ processors, and up to $p$ positions in this memory may be simultaneously read or written
  in one step\cite{shiloach1981}.  As with Roughsort, a number of algorithms take advantage of this model\textquoteright s 
  assumptions to provide performance improvements over their serial counterparts\cite{shiloach1982}\cite{raj1989}.

  Conspicuous, however, with these and other CRCW PRAM algorithms is a lack of presented implementations.  The model is merely
  that, and thus implementing such an algorithm requires a suitable proxy which shares the characteristics of CRCW PRAM.  In 
  contemporary computing, the most notable such proxy is graphics processing unit (GPU) parallelization.

  GPUs invert a core principle of general purpose computing to yield a performance improvement on specific workloads.  Rather
  than having one or a handful of general purpose, powerful multiprocessors, GPUs opt for a different model: many less-powerful,
  special purpose processors.  Contemporary examples include the Skylake Intel Xeon E3-1275 v5 CPU and the Nvidia GTX 1080 
  GPU\footnote{Both were used on this project, for the evaluation of sequential and parallel implementations of these algorithms,
  respectively.}. The Xeon has 4 processor cores whereas the Nvidia card has 2560 cores \cite{xeon} \cite{geforce}.

  Nvidia\textemdash the manufacturer of the aforementioned graphics card\textemdash makes available an API suite and SDK which
  allows developers to readily access the GPU: CUDA.  This platform provides developers a simple interface with which they can
  offload processing of CUDA-augmented C\texttt{++} software payloads to CUDA-supported Nvidia cards.

  The parallelism provided by Nvidia and CUDA in this scheme can be realized by developers who develop their payloads in specific
  ways.  Fundamentally, programs running on an Nvidia card via CUDA operate differently than their  CPU counterparts.  In general,
  a line of code executing on an Nvidia card is not processed in isolation.  Rather, the Nvidia CUDA scheme provides for single 
  instruction, multiple thread (SIMT) execution, as opposed to the usual single instruction, single data scheme (SISD) provided by
  general purpose CPUs.  In contrast to a single instruction accessing or updating one memory location as the result of one 
  instruction executing on a single core of a general purpose CPU, a single instruction executing on an Nvidia card is executing 
  inside multiple threads on multiple respective cores simultaneously, each of which operates on a different location in memory.

  When a developer wants to take advantage of this parallelism, they structure their code such that the multiple results of the same
  instruction executing with values from different, largely contiguous memory locations is a useful component of the overall 
  program's payload.  Contiguity is an important caveat, due to the way memory accesses in Nvidia cards are structured.  Effective
  memory bandwidth is significantly reduced when memory accesses are not grouped in such a manner as to take advantage of the 
  architecture, reducing memory bandwidth by over 90\% on even higher-end Nvidia cards \cite{coalesce}. Memory access which complies
  with this constraint is said to have \textit{coalescence}.

  The core component of an Nvidia payload is a thread; an instance of the series of instructions in the overall CUDA payload.
  Multiple threads run in a group called a block.  Blocks contain a developer-specified number of threads each.  Between threads and
  blocks is the concept of a warp.  Warps are collections of 32 threads executing simultaneously on separate processor cores of the
  Nvidia card.  Usually, all threads in a warp are executing the same instruction simultaneously. However, in situations where a 
  conditional is evaluated by members of a warp, all threads for whom the conditional yields true will execute, and subsequently all
  threads with a false conditional result will execute (or vice versa), in a phenomenon called \textit{warp divergence}.  This yields
  from the SIMT architecture, where all cores on which a warp executes must be processing the same instruction or temporarily 
  switched off and not processing at all.  Blocks are divided up into multiple warps for scheduling and execution on an Nvidia GPU.
  Collections of these blocks are referred to as a grid.  Finally, encompassing all of these components is the kernel.  Kernels are 
  the series of instructions which threads, warps, blocks, and grids comprise an instance of and represent the sequence of code to 
  actually be transferred to and executed on GPU.

  Nvidia provides a simple mechanism by which developers can access the parallelism provided by CUDA.  Each of the above divisions of 
  the CUDA payload are provided with an identifier.  A tuple composed of the thread, block, and grid identifiers\footnote{Concerns with
  warps are abstracted away from the developer.  Consequently warp identifiers are not needed and are not available.} uniquely 
  identifies that component of the workload.  Using this information, individual threads all executing the same instruction (such as a
  memory read from an array) can perform that instruction on distinct data (access different memory locations in the array) \cite{cuda}.

@@

  In achieving parallel radius finding, Altman and Chlebus noted that the $LR$, $RL$, and $DM$ sequences necessary to determine the 
  radius $k$ of a roughsorted sequence could be computed in $O(\lg k)$, while $k$ is again at most $n - 1$ \cite{altman90}, with the 
  above underlying assumptions.

  Using CUDA as a proxy for CRCW PRAM, its techniques provided a platform to develop the GPU parallel versions of sequence radius finding
  and subsequent roughsorting.

  In support of the above discussed parallel Roughsort implementation, CUDA implementations of the $RL$, $LR$, and $DM$ sequence determination
  algorithms were developed.  This was done by assigning each thread a unique identifier and then, analogous to Altman and Chlebus\textquoteright
  work, having each thread manage the corresponding slot in the global unsorted array \cite{altman89}.  Each thread performed the parallel min- 
  then max-prefix operations until the respective lists were efficiently determined ordered.  Sorting was determined by having each thread 
  determine if its assigned slot in the $LR$ or $RL$ list was greater than the next element in the list, and having each thread globally assert
  any true (unsorted) result to global memory, available to all other threads.

  Once the $RL$ and $LR$ lists were completed, the $DM$ list was computed by using Altman and Chlebus\textquoteright s serial radius techniques,
  while using the $O(\lg k)$ bounded techniques introduced by their $RL$ and $LR$ sequence determination algorithms.  However, a notable divergence
  from the serial version of the serial $DM$ list algorithm was taken.  The inner loop of the serial $DM$ list algorithm added a criteria which 
  assured that $LR[i]$ (the $i$th element of the $LR$ list) was always greater than or equal to $RL[i]$ (the $i$th element of the $RL$ list).  
  Only the power of 2 greater than the actual value of $k$ is needed for parallel Roughsort to function, so this criteria was removed in the 
  interest of efficiency and simplicity\cite{altman89}.

  Finally, to determine the radius from the $DM$ list, the maximum element needed to be found.  For this purpose, the \textit{Thrust} CUDA library
  was used, which provides CUDA-optimized implementations of a number of C\texttt{++} STL functions\cite{thrust}.  One such STL function is the 
  \texttt{max\textunderscore element} function.  Using the \textit{Thrust} implementation of this function, maximum elements of large arrays can 
  be found expediently on Nvidia hardware using CUDA\cite{extrema}.  Using this function yielded the maximum element of the $DM$ list, which is 
  \textit{k}.  Further, to ensure all values quiesced during $LR$, $RL$, and $DM$ list computations, a separate kernel was launched to perform 
  the final $DM$ computations.

@@

  Running these kernels yielded disappointing results, however.  Figures 1 and 2 compare the runtime of the sequential algorithm to the parallel.
  For random lists composed together in such a manner as to yield a specific $k$ value for the list (as discussed later in the paper)
  \footnote{Lists which had a specific \textit{k} value were generated ahead of time and then evaluated using the radius finders to prove both
  correctness and provide for validity in this experimentation}, sequential CPU computations exceeded performance on GPU for any tested $n$ when
  values of $k$ exceeded 6, as see in Figure 2\footnote{These results reversed when performed in a Windows environment, compiled with Visual 
  Studio 2015 Community and run on commodity hardware.  The authors suspect this to be a false result and posit that optimization differences on
  the C\texttt{++} STL \texttt{nth\textunderscore element} function are the cause.  This invites further investigation, however.}.  Figure 1 
  reports more positive results for $k$ equal to 2, showing that parallel run times were equalled or exceeded by sequential runtimes for tested 
  values $n$ equal to or in excess of 150,000, though this result was at best constant factor improvement over the sequential implementation and
  does not demonstrate the expected $O(\lg k)$ runtime, emphasizing a more fundamental issue with the platform, implementation, or both.

@@

Regrettably, \texttt{Thrust} has yet to provide an \texttt{nth_element} selection algorithm implementation, so we replaced it with calls
  to \texttt{Thrust}'s \texttt{sort} algorithm, which performs a ``state-of-the-art'', highly-optimized radix sort
  implementation when invoked over arrays of primitive integer elements\cite{thrustradix, leischner09}.
Radix sort is a generalized counting sort that runs in linear time when operating over fixed-size array elements.
While radix sort requires additional storage (unlike Introselect, which was used in the sequential implementation), this
  additional storage ought to be acceptable as the segments we wish to partition ought to be relatively small.

Each \texttt{Thrust} \texttt{sort} call issues at least one kernel launch; we attempted to run radix sorts over the entire array in
  parallel by establishing a fixed number of CUDA \textit{streams}.
Whereas kernels launched through the same stream are executed in order, up to 16 kernels launched through distinct streams can be
  run concurrently on the GPU, available resources permitting\cite{cuda}.
We established a fixed number of CUDA streams and launched each \texttt{sort} call through them in a circular, round-robin
  fashion.

@@

We implemented this approach using a CUDA kernel where each thread in the launch is responsible for sorting exactly one segment
  from step 1 or 2; the steps are divided into separate kernel launches, each given an offset from which begin segmenting the
  array.
While the threads in each launch run concurrently\footnote{For large arrays, the GPU can't host every thread simultaneously and
  must sequentially execute the thread blocks in batches.}, each thread sorts its designated segment sequentially.
In contrast to \supercite{altman90}, we used \texttt{Thrust}'s radix sort instead of its Mergesort implementation in order to leverage
  the linear runtime of the former.
While this approach to the parallel Roughsort was far more successful than the previous, it still struggled against
  its competitors (see Section 9).

@@

\section{Implementation Performance Analysis}
We tested our sequential and parallel Roughsort implementations against sequential and parallel Mergesort (as implemented by
  the C\texttt{++} STL and by CUDA \texttt{Thrust}, respectively) as well as against our simple, sequential Bubblesort implementation.
We chose Mergesort since it's a classic sort algorithm that also scales very well with parallel execution, capable
  of achieving a speedup of at least $\Theta(n / \lg^2 n)$ when given unlimited processors for a runtime of
  $\Theta(\lg^3 n)$\cite{clrs}.
While Bubblesort has a quadratic runtime upper bound, its lightweight, linear lower bound emerges when given nearly-sorted
  sequences and provides a means of comparing our implementation's performance for small values of $k$.

@@

Figures 5 and 6 fix $n$ and vary $k$, yielding similar results.
For arrays of these lengths, the parallel Roughsort's crossover point lies somewhere around the interval $20 \leq k \leq 30$,
  where it manages to best the three sequential algorithms.
An invariant among all four graphs is the utter dominance of the parallel Mergesort implementation provided by \textit{Thrust},
  whose runtimes never depart from the $x$-axis.

@@

\clearpage
\section{Explanation of Results}
  The source papers theorized that parallel radius finding would run at a $O(\lg(k))$ bound, so these contrary result must then be explained.  
  There are several possible reasons for this which should be considered, as there are similarities but also differences between GPU and the 
  assumed CRCW PRAM.
    Given many GPU processor threads may update multiple memory locations simultaneously, and likewise many memory locations may be read in 
  parallel on GPU, the underlying NVIDIA hardware that CUDA provides access to is reminiscent of CRCW PRAM.  However, GPUs differ from true
  CRCW PRAM in fundamental ways.  First, GPU memory is finite, while PRAM is of arbitrary size.  Furthermore, for GPU memory access times to
  provide the concurrent read and write characteristics reminiscent of PRAM, these memory operations must be performed on contiguous portions
  of GPU memory in a \textit{coalescing manner}, whereas CRCW PRAM has no such ordering limitation.  Foremost in the differences between CRCW
  PRAM and GPU architectures however is the warp divergence concept addressed above.  What this implies for memory operations is that workloads
  exist which may prevent all but one thread in a warp from performing a memory operation at a particular time.  Consequently, there are
  practical limitations to what a GPU may achieve, and these limitations restrict the workloads which are practical to run on a GPU.  Not 
  only must a GPU workload constrain its memory usage within practical limitations, the accesses to that memory must be orderly, and the 
  underlying algorithm performing those accesses must be amenable to parallelization on GPU if the advantages of GPU processing over CPU 
  processing are to be realized\cite{dehne2010exploring}.
  Likewise, while the parallel Roughsort results were more in line with expectations, that it failed to best the heavily optimized \texttt{Thrust}
  Mergesort implementation confirms that issues impacting parallel radius finding similarly affect parallel Roughsort's performance.

\section{Conclusion and Further Research}
  Ultimately, results for this project did not result in a parallel Roughsort implementation which convincingly bested other comparison sort 
  algorithms. Most problematic was the linear performance of the parallel radius finder, which was expected to run in logarithmic time.  Roughsort
  results were encouraging, but ultimately parallel Mergesort dominated on all tested values of $k$, demonstrating the impracticality of the
  parallel Roughsort implementation for workloads which would clearly benefit more from parallel Mergesort.  The authors posit that this is 
  largely a consequence of the differences between CRCW PRAM and the practical limitations of Nvidia GPU architectures.  
  In the same vein, sequential Roughsort was bested in the domain of low $k$ values by Bubblesort, due to Bubblesort's expected good performance
  on nearly sorted sequences, closing off yet another avenue in which Roughsort might have utility.

  However, as a first implementation, this contributes to the body of work and provides several future avenues for exploration and possible 
  improvement over these results.
  The practical considerations of the GPU hardware were not explicitly addressed while architecting and designing parallel implementations of
  the mentioned algorithms.  The dominating \texttt{Thrust} Mergesort implementation was once in the same situation, where additional optimizations
  were discovered and then implemented.  Work may now be done to coalesce memory access requests across threads in the aforementioned payloads.
  Likewise, optimizations which might have the algorithms take advantage of more performant thread-local memory, specify beneficial cache 
  parameters, avoid warp divergence, and generally address practical constraints of the GPU can now be investigated \cite{cuda}.  Further 
  exploration may also be undertaken into CUDA implementations of CRCW parallel algorithms which might address the limitations seen in these 
  results.  For instance, an implementation of Cole's parallel median algorithm might enable a more efficient parallel Roughsort\cite{cole85}.

  Finally, this body of work introduces a useful concept: the $k$ value.  Not only does determining the radius of a sequence provide for the 
  possibility of selecting the best sorting algorithm for the situation, being able to generate sequences with an arbitrary $k$ value is useful
  in the testing and evaluation of sorting algorithms, as was done here.  Not only could this work be cumulatively used to implement sorting 
  techniques which select the optimal algorithm for the situation based on the $k$ value, evaluation of sorting algorithms becomes more scientific
  and thus effective.
